Headgroup Structure and Cation Binding in Phosphatidylserine Lipid Bilayers

### 2019-09-11 Manuscript/manuscriptPS.tex
Molecular electrometer

\subsection{Using the molecular electrometer concept to compare ion binding to negatively charged lipid bilayers 
in simulations and in experiments}

The $S_{\rm CH}$ of the $\alpha$ and $\beta$ carbons in the PC headgroup
decrease proportionally to the amount of positive
charge bound to the bilayer \cite{akutsu81,altenbach84,seelig87},
and can therefore be used to measure the ion binding affinity.
In addition to ions, the correlation between bound charge and headgroup order parameter change
is empirically observed also for peptides, charged amphiphiles, local anesthetics and charged lipids~\cite{scherer87,beschiaschvili91}.
This concept, known as the molecular electrometer, is especially useful for 
comparison between simulations and experiments, as
the headgroup $S_{\rm CH}$ at varying cation
concentrations can be easily calculated from
simulations~\cite{catte16}. The headgroup $S_{\rm CH}$
of negatively charged PS and PG lipids also exhibit systematic --- although less well understood ---dependencies on the bound charge~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, measuring the PC headgroup $S_{\rm CH}$ from 
mixed (here PS:PC) bilayers~\cite{roux86,roux90,roux91} (see also SI section \ref{electrometerFORmixtures}) provides a more straightforward way of characterizing the ion binding to negatively charged membranes.

Calibrating the PC $S_{\rm CH}$ response to a known amount of bound charge~\cite{catte16,melcr18} is an important preliminary step for using the molecular electrometer.
This can be done using experimental data from mixtures of PC and
monovalent cationic surfactants (such as POPC and dihexadecyldimethylammonium,
see SI section \ref{electrometerCALIBRATION})~\cite{scherer89,melcr18}.
Additionally, the response of PC headgroup $S_{\rm CH}$ to the negatively
charged PS follows the molecular electrometer in experiments~\cite{scherer87},
which we also quantify here (see SI section \ref{electrometerFORmixtures}).

Studies applying the molecular electrometer have used two different definitions for salt concentration:
The concentrations can be reported either before \cite{akutsu81,roux90,catte16}, or after \cite{altenbach84,melcr18} solvating the lipids.
In the former case, binding of ions to the lipids leads to a lower bulk concentration than in what was present in the original solvent.
However, the choice of definition has only a marginal effect
to the results in simulations with realistic ion binding affinity
(see SI section \ref{concentrationDEFsection}).
In this work we use the former definition (concentration before solvating the lipids) to be consistent with the reference
experimental data \cite{roux90}.

@@

\begin{acknowledgement}
  HA acknowledges the Osk. Huttunen Foundation and the Finnish Academy of Science and Letters (Foundations' Post Doc Pool) for financial support.
  FFR acknowledges Tecnol\'{o}gico Nacional de M\'{e}xico Campus Zacatecas Occidente and
  Direcci\'{o}n General de Asuntos del Personal Acad\'{e}mico (DGAPA)
  Programa de Apoyo a Proyectos de Investigaci\'{o}n e Innovaci\'{o}n
  Tecnol\'{o}gica (PAPIIT) IG100416 for financial support and Cl\'{u}ster
  H\'{i}brido de Superc\'{o}mputo Xiuhcoatl-CINVESTAV and Miztli-Direcci\'{o}n de
  C\'{o}mputo y de Tecnolog\'{i}as de Informaci\'{o}n y Comunicaci\'{o}n (DGTIC)-Universidad Nacional Aut\'{o}noma de M\'{e}xico (UNAM) (Project
  LANCAD-UNAM-DGTIC-028) facilities for computing-time allocation.
  MJ acknowledges financial support from the Emil Aaltonen Foundation
  and CSC -- IT Center for Science for computational resources.
  JJM gratefully acknowledges financial support from the Carlsberg Foundation in the form of a postdoctoral fellowship while at the University of Chicago (grants CF15-0552, CF16-0639, and CF17-0783) and the research framework provided by the Research Computing Center at the University of Chicago.
  OHSO acknowledges financial support from Academy of Finland (315596),
  Integrated Structural Biology Research Infrastructure of
  Helsinki Institute of Life Science (Instruct-HiLIFE), and
  CSC -- IT Center for Science for computational resources.
  TJP wishes to acknowledge the use of the Iridis computing resources at the University of Southampton.
\end{acknowledgement}

>>>>

\subsection{Using the molecular electrometer to compare ion binding to negatively charged lipid bilayers 
in simulations and in experiments}

The $S_{\rm CH}$ of the $\alpha$ and $\beta$ carbons in the PC headgroup
decrease proportionally to the amount of positive
charge bound to the bilayer \cite{akutsu81,altenbach84,seelig87},
and can therefore be used to measure the ion binding affinity.
In addition to ions, the correlation between bound charge and headgroup order parameter change
is empirically observed also for peptides, charged amphiphiles, local anesthetics and charged lipids~\cite{scherer87,beschiaschvili91}.
This concept, known as the molecular electrometer, is especially useful for 
comparison between simulations and experiments, as
the headgroup $S_{\rm CH}$ at varying cation
concentrations can be easily calculated from
simulations~\cite{catte16}. The headgroup $S_{\rm CH}$
of negatively charged PS and PG lipids also exhibit systematic dependencies on the bound charge, but these are less well understood than for PC~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, measuring the PC headgroup $S_{\rm CH}$ in 
mixed (here PS:PC) bilayers~\cite{roux86,roux90,roux91} (see also Supporting Information section \ref{electrometerFORmixtures}) provides a more straightforward way of characterizing ion binding to such negatively charged membranes.

Calibrating the PC headgroup $S_{\rm CH}$ response to a known amount of bound charge~\cite{catte16,melcr18} is an important preliminary step for using the molecular electrometer.
This can be done using experimental data from mixtures of PC and
monovalent cationic surfactants (such as POPC and dihexadecyldimethylammonium,
see SI section \ref{electrometerCALIBRATION})~\cite{scherer89,melcr18}.
Additionally, the response of PC headgroup $S_{\rm CH}$ to the negatively
charged PS follows the molecular electrometer in experiments~\cite{scherer87},
which we also quantify here (see SI section \ref{electrometerFORmixtures}).

Studies applying the molecular electrometer have used two different definitions for salt concentration:
The concentrations reported either before \cite{akutsu81,roux90,catte16}, or after \cite{altenbach84,melcr18} solvating the lipids.
In the former case, binding of ions to the lipids leads to a lower bulk concentration than what was present in the original solvent.
However, the choice of definition has only a marginal effect
to the results in simulations with realistic ion binding affinity
(see SI section \ref{concentrationDEFsection}).
In this work, we use the concentration before solvating the lipids to be consistent with the reference
experimental data \cite{roux90}.

@@

\begin{acknowledgement}
  HA acknowledges the Osk. Huttunen Foundation and the Finnish Academy of Science and Letters (Foundations' Post Doc Pool) for financial support.
  FFR acknowledges Tecnol\'{o}gico Nacional de M\'{e}xico Campus Zacatecas Occidente and
  Direcci\'{o}n General de Asuntos del Personal Acad\'{e}mico (DGAPA)
  Programa de Apoyo a Proyectos de Investigaci\'{o}n e Innovaci\'{o}n
  Tecnol\'{o}gica (PAPIIT) IG100416 for financial support and Cl\'{u}ster
  H\'{i}brido de Superc\'{o}mputo Xiuhcoatl-CINVESTAV and Miztli-Direcci\'{o}n de
  C\'{o}mputo y de Tecnolog\'{i}as de Informaci\'{o}n y Comunicaci\'{o}n (DGTIC)-Universidad Nacional Aut\'{o}noma de M\'{e}xico (UNAM) (Project
  LANCAD-UNAM-DGTIC-028) facilities for computing-time allocation.
  MJ acknowledges financial support from the Emil Aaltonen Foundation
  and CSC -- IT Center for Science for computational resources.
  JJM gratefully acknowledges financial support from the Carlsberg Foundation in the form of a postdoctoral fellowship while at the University of Chicago (grants CF15-0552, CF16-0639, and CF17-0783) and the research framework provided by the Research Computing Center at the University of Chicago.
  OHSO acknowledges financial support from Academy of Finland (315596),
  Integrated Structural Biology Research Infrastructure of
  Helsinki Institute of Life Science (Instruct-HiLIFE), and
  CSC -- IT Center for Science for computational resources.
  TJP wishes to acknowledge the use of the Iridis computing resources at the University of Southampton. BK gratefully acknowledges financial support from the International Max Planck Research School (IMPRS) on Multiscale Bio-Systems. 
\end{acknowledgement}
